Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-06-18 paper_jasa.tex
UMP replaced by conventional

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. Until recently there were no formal visual methods in place for determining statistical significance of findings. This changed, when \citet{buja:2009} conceptually introduced two protocols for formal tests of visual findings. In this paper we take this a step further by comparing the lineup protocol \citep{buja:2009} against classical statistical testing  of the significance of regression model parameters. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides results comparable to the uniformly most powerful (UMP) test, out-performs the UMP when data is contaminated, and, there may be some super-individuals who yield better power than the UMP test even in the most difficult trials.
\end {abstract}

@@

The paper is organized as follows. Section \ref{sec:visual_test} defines terms as used in visual inference, and describes how to estimate the important quantities from experimental data. The effect of the lineup size and number of observers on the power of the test is discussed in Section \ref{sec:size}. Section \ref{sec:regression} focuses on the application of  visual inference to linear models.  Section \ref{sec:simulation} describes the results of the three simulation experiments conducted to compare the power of the lineup protocol with the equivalent uniformly most powerful (UMP) and an analysis of the resulting data. The last section describes the next steps in this research.

@@

The classical test in this scenario uses $T= \hat{\beta_k}/ se(\hat{\beta_k})$ and rejects the null hypothesis if $T$ is extreme on the scale of a $t$ distribution with $n-p$ degrees of freedom. This is the uniformly most powerful (UMP) test \green{reference needed here} and it forms the benchmark upon which we evaluate the visual test. (This is also the UMP test for the continuous covariate, case 2.) To calculate what we might expect for the power of the visual test, under perfect conditions, first assume that the observer is able to pick the plot with the smallest $p$-value from a lineup plot.  This leads to the decision to reject $H_0$ when $p_{D} < p_0$, where $p_{D}$ is the classical $p$-value as details given in Lemma \ref{lemma}. Thus the expected power of the visual test in this scenario is

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the UMP test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. Notice that the expected power of the visual test is almost as good as the power of UMP test. \green{What do you mean by this? Where does this ``fact'' come from?} In fact, the only loss we experience stems from comparing the test statistic to only a discrete number of  $m-1$ samples from the null distribution.

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% Equation \ref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \ref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.95}{\includegraphics{power_expected.pdf}}
       \caption{Expected power of visual test  and the power of UMP test for sample size $n =100$, $\sigma = 12$ and $m$=20. }
       \label{fig:power_expected}
\end{figure}

@@

The experiment is designed to study the ability of human observers to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated a range of values of $\beta_2~ (=0, 1, 3, 5, 7, 8, 10, 16)$, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical UMP test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making a side-by-side boxplot of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and these were plotted in the same way as side-by-side boxplots. The actual data plot was randomly placed among these null data plots to produce the lineup. Figure \ref{fig:test_category} is an example of one of these lineups. The actual data plot is number 11 which was generated for $n$=300, $\beta_2$=15 and $\sigma$=12.

@@

The number of evaluations required for each lineup to provide reasonable estimates of the power is determined by the variance of the estimated power. Suppose $\gamma$ denotes the classical UMP test power for each parameter combination shown in table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of classical UMP test (Figure \ref{fig:power_expected}) the proportion of successful evaluations of lineups should give a rough estimate for  $\gamma$. For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \frac 1 n_{\gamma} \gamma(1-\gamma)} \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$

@@

The first two simulation experiments use data generated with a normal error model, satisfying the conditions for uniformly most powerful (UMP) test procedures. In these situations there exists a test, and there would, in general, be no need to use visual inference. The simulation is conducted with the hopes that the visual test procedure, although not expected to beat the UMP, will compare favorably with it. This third simulation is closer to the mark for the purpose of visual inference. The assumptions for the UMP are violated, by contaminating the data. The contamination makes the estimated slopes effectively 0, yet the true value of slope parameter is not.  The data is generated from the following model:

@@

We fit model \ref{mixed} with effect $E=\beta*n/\sqrt(n)$ as the only fixed effect covariate without intercept. We also set $offset = log(0.05/95)$ so that the estimated power has a fixed lower limit = 0.05 (Type-I error) for effect $E=0$. The power curves obtained by fitting model \ref{mixed} to the data for all the experiments applying different screening criteria is shown in figure \ref{fig:power_screening}. This shows how the screening criteria may affect the results. The corresponding power curves of UMP test are also shown. Notice that no mater what criteria we apply the result does not change much for experiment 2. Also, estimated power curve with criteria 5 for experiment 3 is above the UMP test power curve.

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{power_screening.pdf}}
       \caption{The overall power estimated from model \ref{mixed} are shown for various data sets obtained by applying 6 screening criteria.  The corresponding power curve for Uniformly Most powerful (UMP) test is shown in black solid lines for comparison.}
       \label{fig:power_screening}
\end{figure}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.45}{\includegraphics{effect_power_exp1.pdf}}
       \scalebox{0.45}{\includegraphics{effect_power_exp2.pdf}}
       \caption{Observed power shown by loess smoother and the power of UMP test for different effect size $E$ as per experiment 1 and 2.}
       \label{fig:power_effect1}
\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.4}{\includegraphics{power_loess_exp3.pdf}}
       \caption{Observed power shown by loess smoother with simultaneous bootstrap confidence band and the power of UMP test for sample size $n= 100$ and $\sigma = 5$ as per experiment 3. Notice that for experiment 3, power does not seem to depend on the parameter $\beta$. Di/Heike, I am not sure how I can relate this to demonstrate the power of visual inference when normal conditions are not met. Does this really test our hypothesis $H_0: \beta=0$ or some other hypothesis about the existence of cluster in the data?}
       \label{fig:power_loess3}
\end{figure*}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.60}{\includegraphics{power_screening_subject.pdf}}
      \caption{The subject specific power estimated from model \ref{mixed} are shown for various data sets obtained by applying 6 screening criteria.  The corresponding power curve for Uniformly Most powerful (UMP) test is shown in pink lines for comparison.}
       \label{fig:power_screening_subject}	
\end{figure}

@@

\green{third experiment doesn't really have a p-value, does it? Need to explain this. Why are there still so many small p-values in experiment 3?}

@@

\begin {abstract} 
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. Until recently there were no formal visual methods in place for determining statistical significance of findings. This changed, when Buja et al.(2009) conceptually introduced two protocols for formal tests of visual findings. In this paper we take this a step further by comparing the lineup protocol (Buja et al.2009) against classical statistical testing  of the significance of regression model parameters. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides results equivalent to the uniformly most powerful (UMP) test and for some scenarios yields better power than the UMP test.
\end {abstract}

>>>>

\begin {abstract}  
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. Until recently there were no formal visual methods in place for determining statistical significance of findings. This changed, when \citet{buja:2009} conceptually introduced two protocols for formal tests of visual findings. In this paper we take this a step further by comparing the lineup protocol \citep{buja:2009} against classical statistical testing  of the significance of regression model parameters. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides results comparable to the conventional tests, out-performs them when data is contaminated, and, there may be some super-individuals who yield better power than the conventional test even in the most difficult trials.
\end {abstract}

@@

The paper is organized as follows. Section \ref{sec:visual_test} defines terms as used in visual inference, and describes how to estimate the important quantities from experimental data. The effect of the lineup size and number of observers on the power of the test is discussed in Section \ref{sec:size}. Section \ref{sec:regression} focuses on the application of  visual inference to linear models.  Section \ref{sec:simulation} describes the results of the three simulation experiments conducted to compare the power of the lineup protocol with the equivalent conventional test and an analysis of the resulting data. The last section describes the next steps in this research.

@@

The conventional test in this scenario uses $T= \hat{\beta_k}/ se(\hat{\beta_k})$ and rejects the null hypothesis if $T$ is extreme on the scale of a $t$ distribution with $n-p$ degrees of freedom.  It forms the benchmark upon which we evaluate the visual test. To calculate what we might expect for the power of the visual test, under perfect conditions, first assume that the observer is able to pick the plot with the smallest $p$-value from a lineup plot.  This leads to the decision to reject $H_0$ when $p_{D} < p_0$, where $p_{D}$ is the classical $p$-value as details given in Lemma \ref{lemma}. Thus the expected power of the visual test in this scenario is

\begin{equation}\label{power_exp} 
   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test obtained from Equation \ref{power_exp}. Notice that the expected power of the visual test is almost as good as the power of conventional test. \green{What do you mean by this? Where does this ``fact'' come from?} In fact, the only loss we experience stems from comparing the test statistic to only a discrete number of  $m-1$ samples from the null distribution.

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% Equation \ref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \ref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.95}{\includegraphics{power_expected.pdf}}
       \caption{Expected power of visual test  and the power of conventional test for sample size $n =100$, $\sigma = 12$ and $m$=20. }
       \label{fig:power_expected}
\end{figure}

@@

The experiment is designed to study the ability of human observers to detect the effect of a single categorical variable $X_2$ (corresponding to parameter $\beta_2$) in a two variable ($p=2$) regression model (Equation \ref{multi}). Data is simulated a range of values of $\beta_2~ (=0, 1, 3, 5, 7, 8, 10, 16)$, two different sample sizes ($n=100, 300$) and two standard deviations of the error ($\sigma=5, 12$). The range of $\beta_2$ values was chosen so that estimates of the power would produce reasonably continuous power curves, comparable to that calculated for the theoretical conventional test. Values were fixed for other regression parameters, $\beta_0 = 5$,  $\beta_1=15$, and the values for $X_1$ were randomly generated from a Poisson $(\lambda=30)$ distribution, which is almost Gaussian. Data sets were generated with parameter values shown in Table \ref{tbl:experiment_params}. Three replicates at each level were generated. These produced 60 different ``actual data sets'', and thus, 60 different lineups. To produce the lineups, the null model was fit to each actual data set, to obtain residuals and parameter estimates. The actual data plot was generated by making a side-by-side boxplot of the residuals (Table \ref{tbl:stat_multiple}, case 3). The 19 null data sets were generated by simulating from  $N(0, {\hat{\sigma}}^2)$, and these were plotted in the same way as side-by-side boxplots. The actual data plot was randomly placed among these null data plots to produce the lineup. Figure \ref{fig:test_category} is an example of one of these lineups. The actual data plot is number 11 which was generated for $n$=300, $\beta_2$=15 and $\sigma$=12.

@@

The number of evaluations required for each lineup to provide reasonable estimates of the power is determined by the variance of the estimated power. Suppose $\gamma$ denotes the conventional test power for each parameter combination shown in table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of conventional test (Figure \ref{fig:power_expected}) the proportion of successful evaluations of lineups should give a rough estimate for  $\gamma$. For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \frac 1 n_{\gamma} \gamma(1-\gamma)} \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$

@@

The first two simulation experiments use data generated with a normal error model, satisfying the conditions for conventional test procedures. In these situations there exists a test, and there would, in general, be no need to use visual inference. The simulation is conducted with the hopes that the visual test procedure, although not expected to beat the conventional test, will compare favorably with it. This third simulation is closer to the mark for the purpose of visual inference. The assumptions for the conventional test are violated, by contaminating the data. The contamination makes the estimated slopes effectively 0, yet the true value of slope parameter is not.  The data is generated from the following model:

@@

We fit model \ref{mixed} with effect $E=\beta*n/\sqrt(n)$ as the only fixed effect covariate without intercept. We also set $offset = log(0.05/95)$ so that the estimated power has a fixed lower limit = 0.05 (Type-I error) for effect $E=0$. The power curves obtained by fitting model \ref{mixed} to the data for all the experiments applying different screening criteria is shown in figure \ref{fig:power_screening}. This shows how the screening criteria may affect the results. The corresponding power curves of conventional test are also shown. Notice that no mater what criteria we apply the result does not change much for experiment 2. Also, estimated power curve with criteria 5 for experiment 3 is above the conventional test power curve.

\begin{figure}[hbtp]
   \centering
       \scalebox{0.70}{\includegraphics{power_screening.pdf}}
       \caption{The overall power estimated from model \ref{mixed} are shown for various data sets obtained by applying 6 screening criteria.  The corresponding power curve for conventional test is shown in black solid lines for comparison.}
       \label{fig:power_screening}
\end{figure}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.45}{\includegraphics{effect_power_exp1.pdf}}
       \scalebox{0.45}{\includegraphics{effect_power_exp2.pdf}}
       \caption{Observed power shown by loess smoother and the power of conventional test for different effect size $E$ as per experiment 1 and 2.}
       \label{fig:power_effect1}
\end{figure*}

@@

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.4}{\includegraphics{power_loess_exp3.pdf}}
       \caption{Observed power shown by loess smoother with simultaneous bootstrap confidence band and the power of conventional test for sample size $n= 100$ and $\sigma = 5$ as per experiment 3. Notice that for experiment 3, power does not seem to depend on the parameter $\beta$. Di/Heike, I am not sure how I can relate this to demonstrate the power of visual inference when normal conditions are not met. Does this really test our hypothesis $H_0: \beta=0$ or some other hypothesis about the existence of cluster in the data?}
       \label{fig:power_loess3}
\end{figure*}

@@

\begin{figure}[hbtp]
   \centering
       \scalebox{0.60}{\includegraphics{power_screening_subject.pdf}}
      \caption{The subject specific power estimated from model \ref{mixed} are shown for various data sets obtained by applying 6 screening criteria.  The corresponding power curve for conventional test is shown in pink lines for comparison.}
       \label{fig:power_screening_subject}	
\end{figure}

@@

\green{third experiment doesn't really have a p-value, does it? Need to explain this. Why are there still so many small p-values in experiment 3?} \blue{$p$-values are computed fitting the full model to the contaminated data anyway.}

@@

\begin {abstract} 
Statistical graphics play a crucial role in exploratory data analysis, model checking and diagnosis. Until recently there were no formal visual methods in place for determining statistical significance of findings. This changed, when Buja et al.(2009) conceptually introduced two protocols for formal tests of visual findings. In this paper we take this a step further by comparing the lineup protocol (Buja et al.2009) against classical statistical testing  of the significance of regression model parameters. A human subjects experiment is conducted using simulated data to provide controlled conditions. Results suggest that the lineup protocol provides results equivalent to the conventional test and for some scenarios yields better power than the conventional test.
\end {abstract}
